### 2019-01-14 submission/manuscript.tex
Render manuscript
Fix legend placement

\section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
High Fidelity DNA
Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}

\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
errors when sequencing 16S rRNA genes from complex
communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}

@@

@@

\linenumbers

\subsection{Abstract}\label{abstract}

\subsection{Introduction}\label{introduction}

@@

\subsection{Results}\label{results}

@@

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g. \emph{L. fermentum}), while
others became more biased (e.g. \emph{E. faecalis}), and others had
little change (e.g. \emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
PCR. Given the subtle nature of the variation in the relative abundances
of each 16S rRNA gene fragment, it was not possible to create
generalizable rules that would explain the bias and the previous
observation regarding the guanine and cytosine relative abundance was
not observed with our data.

@@

\subsection{Discussion}\label{discussion}

@@

\subsection{Materials \& Methods}\label{materials-methods}

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries
{[}(18); \url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The
number of rounds of PCR used for each sample and polymerase 15 and
increased by 5 rounds up to 35 cycles. Insufficient PCR product was
generated using 15 rounds and has not been included in our analysis.

\textbf{\emph{Library generation and sequencing.}} Each PCR condition
(i.e.~combination of polymerase and number of rounds of PCR) were
replicated four times for the mock community and one time for each fecal
sample. Libraries were generated as described previously {[}(18);
\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
were sequenced using the Illumina MiSeq sequencing platform to generate
paired 250-nt reads.

@@

\subsection{Acknowledgements}\label{acknowledgements}

The authors would like to thank all the study participants in ERIN whose
samples were utilized. We also would like to thank Judy Opp and April
Cockburn for their effort in sequencing the samples as part of the
Microbiome Core Facility at the University of Michigan. Additional
thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
earlier drafts of the manuscript and providing helpful critiques. Salary
support for Marc A. Sze came from the Canadian Institute of Health
Research and NIH grant UL1TR002240. Salary support for Patrick D.
Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-Gilbert2018}{}
1. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2018.
Earth microbiome project and global systems biology. mSystems
\textbf{3}.
doi:\href{https://doi.org/10.1128/msystems.00217-17}{10.1128/msystems.00217-17}.

\hypertarget{ref-HMP2012}{}
2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
the healthy human microbiome. Nature \textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Schloss2016}{}
3. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
bacterial census: An update. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.

\hypertarget{ref-Luo2016}{}
4. \textbf{Luo T}, \textbf{Srinivasan U}, \textbf{Ramadugu K},
\textbf{Shedden KA}, \textbf{Neiswanger K}, \textbf{Trumble E},
\textbf{Li JJ}, \textbf{McNeil DW}, \textbf{Crout RJ}, \textbf{Weyant
RJ}, \textbf{Marazita ML}, \textbf{Foxman B}. 2016. Effects of specimen
collection methodologies and storage conditions on the short-term
stability of oral microbiome taxonomy. Applied and Environmental
Microbiology \textbf{82}:5519--5529.
doi:\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.

\hypertarget{ref-Bassis2017}{}
5. \textbf{Bassis CM}, \textbf{Nicholas M. Moore}, \textbf{Lolans K},
\textbf{Seekatz AM}, \textbf{Weinstein RA}, \textbf{Young VB},
\textbf{Hayden MK}. 2017. Comparison of stool versus rectal swab samples
and storage conditions on bacterial community profiles. BMC Microbiology
\textbf{17}.
doi:\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.

\hypertarget{ref-Gorzelak2015}{}
6. \textbf{Gorzelak MA}, \textbf{Gill SK}, \textbf{Tasnim N},
\textbf{Ahmadi-Vand Z}, \textbf{Jay M}, \textbf{Gibson DL}. 2015.
Methods for improving human gut microbiome data by reducing variability
through sample processing and storage of stool. PLOS ONE
\textbf{10}:e0134802.
doi:\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.

\hypertarget{ref-Dominianni2014}{}
7. \textbf{Dominianni C}, \textbf{Wu J}, \textbf{Hayes RB}, \textbf{Ahn
J}. 2014. Comparison of methods for fecal microbiome biospecimen
collection. BMC Microbiology \textbf{14}:103.
doi:\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.

\hypertarget{ref-BautistadelosSantos2016}{}
8. \textbf{Santos QMB-d los}, \textbf{Schroeder JL}, \textbf{Blakemore
O}, \textbf{Moses J}, \textbf{Haffey M}, \textbf{Sloan W}, \textbf{Pinto
AJ}. 2016. The impact of sampling, PCR, and sequencing replication on
discerning changes in drinking water bacterial community over diurnal
time-scales. Water Research \textbf{90}:216--224.
doi:\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.

\hypertarget{ref-Sinha2015}{}
9. \textbf{Sinha R}, \textbf{Chen J}, \textbf{Amir A}, \textbf{Vogtmann
E}, \textbf{Shi J}, \textbf{Inman KS}, \textbf{Flores R},
\textbf{Sampson J}, \textbf{Knight R}, \textbf{Chia N}. 2015. Collecting
fecal samples for microbiome analyses in epidemiology studies. Cancer
Epidemiology Biomarkers \& Prevention \textbf{25}:407--416.
doi:\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.

\hypertarget{ref-Amir2017b}{}
10. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
\textbf{Debelius J}, \textbf{Morton JT}, \textbf{Hyde E},
\textbf{Robbins-Pianka A}, \textbf{Knight R}. 2017. Correcting for
microbial blooms in fecal samples during room-temperature shipping.
mSystems \textbf{2}:e00199--16.
doi:\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.

\hypertarget{ref-Lauber2010}{}
11. \textbf{Lauber CL}, \textbf{Zhou N}, \textbf{Gordon JI},
\textbf{Knight R}, \textbf{Fierer N}. 2010. Effect of storage conditions
on the assessment of bacterial community structure in soil and
human-associated samples. FEMS Microbiology Letters \textbf{307}:80--86.
doi:\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.

\hypertarget{ref-Song2016}{}
12. \textbf{Song SJ}, \textbf{Amir A}, \textbf{Metcalf JL},
\textbf{Amato KR}, \textbf{Xu ZZ}, \textbf{Humphrey G}, \textbf{Knight
R}. 2016. Preservation methods differ in fecal microbiome stability,
affecting suitability for field studies. mSystems \textbf{1}:e00021--16.
doi:\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.

\hypertarget{ref-Costea2017}{}
13. \textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa S},
\textbf{Pelletier E}, \textbf{Alberti A}, \textbf{Levenez F},
\textbf{Tramontano M}, \textbf{Driessen M}, \textbf{Hercog R},
\textbf{Jung F-E}, \textbf{Kultima JR}, \textbf{Hayward MR},
\textbf{Coelho LP}, \textbf{Allen-Vercoe E}, \textbf{Bertrand L},
\textbf{Blaut M}, \textbf{Brown JRM}, \textbf{Carton T},
\textbf{Cools-Portier S}, \textbf{Daigneault M}, \textbf{Derrien M},
\textbf{Druesne A}, \textbf{Vos WM de}, \textbf{Finlay BB},
\textbf{Flint HJ}, \textbf{Guarner F}, \textbf{Hattori M},
\textbf{Heilig H}, \textbf{Luna RA}, \textbf{Hylckama Vlieg J van},
\textbf{Junick J}, \textbf{Klymiuk I}, \textbf{Langella P},
\textbf{Chatelier EL}, \textbf{Mai V}, \textbf{Manichanh C},
\textbf{Martin JC}, \textbf{Mery C}, \textbf{Morita H}, \textbf{O'Toole
PW}, \textbf{Orvain C}, \textbf{Patil KR}, \textbf{Penders J},
\textbf{Persson S}, \textbf{Pons N}, \textbf{Popova M}, \textbf{Salonen
A}, \textbf{Saulnier D}, \textbf{Scott KP}, \textbf{Singh B},
\textbf{Slezak K}, \textbf{Veiga P}, \textbf{Versalovic J}, \textbf{Zhao
L}, \textbf{Zoetendal EG}, \textbf{Ehrlich SD}, \textbf{Dore J},
\textbf{Bork P}. 2017. Towards standards for human fecal sample
processing in metagenomic studies. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.

\hypertarget{ref-Haas2011}{}
14. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
\textbf{Feldgarden M}, \textbf{Ward DV}, \textbf{Giannoukos G},
\textbf{Ciulla D}, \textbf{Tabbaa D}, \textbf{Highlander SK},
\textbf{Sodergren E}, \textbf{Methe B}, \textbf{DeSantis TZ},
\textbf{Petrosino JF}, \textbf{Knight R}, \textbf{and BWB}. 2011.
Chimeric 16S rRNA sequence formation and detection in sanger and
454-pyrosequenced PCR amplicons. Genome Research \textbf{21}:494--504.
doi:\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.

\hypertarget{ref-Sinha2017}{}
15. \textbf{Sinha R}, \textbf{Abu-Ali G}, \textbf{Vogtmann E},
\textbf{Fodor AA}, \textbf{Ren B}, \textbf{Amir A}, \textbf{Schwager E},
\textbf{Crabtree J}, \textbf{Ma S}, \textbf{Abnet CC}, \textbf{Knight
R}, \textbf{White O}, \textbf{Huttenhower C}. 2017. Assessment of
variation in microbial community amplicon sequencing by the microbiome
quality control (MBQC) project consortium. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3981}{10.1038/nbt.3981}.

\hypertarget{ref-Meisel2016}{}
16. \textbf{Meisel JS}, \textbf{Hannigan GD}, \textbf{Tyldsley AS},
\textbf{SanMiguel AJ}, \textbf{Hodkinson BP}, \textbf{Zheng Q},
\textbf{Grice EA}. 2016. Skin microbiome surveys are strongly influenced
by experimental design. Journal of Investigative Dermatology
\textbf{136}:947--956.
doi:\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.

\hypertarget{ref-Caporaso2010}{}
17. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
\textbf{Berg-Lyons D}, \textbf{Lozupone CA}, \textbf{Turnbaugh PJ},
\textbf{Fierer N}, \textbf{Knight R}. 2010. Global patterns of 16S rRNA
diversity at a depth of millions of sequences per sample. Proceedings of
the National Academy of Sciences \textbf{108}:4516--4522.
doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.

\hypertarget{ref-Kozich2013}{}
18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2011}{}
19. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
Reducing the effects of PCR amplification and sequencing artifacts on
16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.

\hypertarget{ref-Bokulich2012}{}
20. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
DA}, \textbf{Caporaso JG}. 2012. Quality-filtering vastly improves
diversity estimates from illumina amplicon sequencing. Nature Methods
\textbf{10}:57--59.
doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.

\hypertarget{ref-Parada2015}{}
21. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
Every base matters: Assessing small subunit rRNA primers for marine
microbiomes with mock communities, time series and global field samples.
Environmental Microbiology \textbf{18}:1403--1414.
doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.

\hypertarget{ref-Wang1996}{}
22. \textbf{Wang GCY}, \textbf{Wang Y}. 1996. The frequency of chimeric
molecules as a consequence of PCR co-amplification of 16S rRNA genes
from different bacterial species. Microbiology \textbf{142}:1107--1114.
doi:\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.

\hypertarget{ref-Potapov2017}{}
23. \textbf{Potapov V}, \textbf{Ong JL}. 2017. Examining sources of
error in PCR by single-molecule sequencing. PLOS ONE
\textbf{12}:e0169774.
doi:\href{https://doi.org/10.1371/journal.pone.0169774}{10.1371/journal.pone.0169774}.

>>>>

\hypertarget{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}{%
\section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
High Fidelity DNA
Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}}

\hypertarget{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}{%
\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
errors when sequencing 16S rRNA genes from complex
communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}}

@@

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure. XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the
Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of
PCR. Given the subtle nature of the variation in the relative abundances
of each 16S rRNA gene fragment, it was not possible to create
generalizable rules that would explain the bias and the previous
observation regarding the guanine and cytosine relative abundance was
not observed with our data.

@@

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

@@

\hypertarget{materials-methods}{%
\subsection{Materials \& Methods}\label{materials-methods}}

@@

\textbf{\emph{PCR protocol.}} Five high fidelity DNA polymerases were
tested including AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA),
KAPA HIFI (Roche, IN, USA), Phusion (New England Biolabs, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
Manufacturer recommendations were followed except for the annealing and
extension times, which were selected based on previously published
protocols (18, 37). Primers targeting the V4 region of the 16S rRNA gene
were used with modifications to generate MiSeq amplicon libraries
{[}@Kozich2013;
\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The number of
rounds of PCR used for each sample and polymerase 15 and increased by 5
rounds up to 35 cycles. Insufficient PCR product was generated using 15
rounds and has not been included in our analysis.

\textbf{\emph{Library generation and sequencing.}} Each PCR condition
(i.e.~combination of polymerase and number of rounds of PCR) were
replicated four times for the mock community and one time for each fecal
sample. Libraries were generated as described previously {[}@Kozich2013;
\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/}{]}. The libraries
were sequenced using the Illumina MiSeq sequencing platform to generate
paired 250-nt reads.

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

The authors thank the study participants in ERIN whose samples were
utilized. We also would like to thank Judy Opp and April Cockburn for
their effort in sequencing the samples as part of the Microbiome Core
Facility at the University of Michigan. Additional thanks to members of
the Schloss lab and Dr.~Marcy Balunas for reading earlier drafts of the
manuscript and providing helpful critiques. Salary support for Marc A.
Sze came from the Canadian Institute of Health Research and NIH grant
UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants
P30DK034933 and 1R01CA215574.

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Gilbert2018}{}%
1. \textbf{Gilbert JA}, \textbf{Jansson JK}, \textbf{Knight R}. 2018.
Earth microbiome project and global systems biology. mSystems
\textbf{3}.
doi:\href{https://doi.org/10.1128/msystems.00217-17}{10.1128/msystems.00217-17}.

\leavevmode\hypertarget{ref-HMP2012}{}%
2. \textbf{Consortium HM}. 2012. Structure, function and diversity of
the healthy human microbiome. Nature \textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\leavevmode\hypertarget{ref-Schloss2016}{}%
3. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
bacterial census: An update. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.

\leavevmode\hypertarget{ref-Luo2016}{}%
4. \textbf{Luo T}, \textbf{Srinivasan U}, \textbf{Ramadugu K},
\textbf{Shedden KA}, \textbf{Neiswanger K}, \textbf{Trumble E},
\textbf{Li JJ}, \textbf{McNeil DW}, \textbf{Crout RJ}, \textbf{Weyant
RJ}, \textbf{Marazita ML}, \textbf{Foxman B}. 2016. Effects of specimen
collection methodologies and storage conditions on the short-term
stability of oral microbiome taxonomy. Applied and Environmental
Microbiology \textbf{82}:5519--5529.
doi:\href{https://doi.org/10.1128/aem.01132-16}{10.1128/aem.01132-16}.

\leavevmode\hypertarget{ref-Bassis2017}{}%
5. \textbf{Bassis CM}, \textbf{Nicholas M. Moore}, \textbf{Lolans K},
\textbf{Seekatz AM}, \textbf{Weinstein RA}, \textbf{Young VB},
\textbf{Hayden MK}. 2017. Comparison of stool versus rectal swab samples
and storage conditions on bacterial community profiles. BMC Microbiology
\textbf{17}.
doi:\href{https://doi.org/10.1186/s12866-017-0983-9}{10.1186/s12866-017-0983-9}.

\leavevmode\hypertarget{ref-Gorzelak2015}{}%
6. \textbf{Gorzelak MA}, \textbf{Gill SK}, \textbf{Tasnim N},
\textbf{Ahmadi-Vand Z}, \textbf{Jay M}, \textbf{Gibson DL}. 2015.
Methods for improving human gut microbiome data by reducing variability
through sample processing and storage of stool. PLOS ONE
\textbf{10}:e0134802.
doi:\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.

\leavevmode\hypertarget{ref-Dominianni2014}{}%
7. \textbf{Dominianni C}, \textbf{Wu J}, \textbf{Hayes RB}, \textbf{Ahn
J}. 2014. Comparison of methods for fecal microbiome biospecimen
collection. BMC Microbiology \textbf{14}:103.
doi:\href{https://doi.org/10.1186/1471-2180-14-103}{10.1186/1471-2180-14-103}.

\leavevmode\hypertarget{ref-BautistadelosSantos2016}{}%
8. \textbf{Santos QMB-d los}, \textbf{Schroeder JL}, \textbf{Blakemore
O}, \textbf{Moses J}, \textbf{Haffey M}, \textbf{Sloan W}, \textbf{Pinto
AJ}. 2016. The impact of sampling, PCR, and sequencing replication on
discerning changes in drinking water bacterial community over diurnal
time-scales. Water Research \textbf{90}:216--224.
doi:\href{https://doi.org/10.1016/j.watres.2015.12.010}{10.1016/j.watres.2015.12.010}.

\leavevmode\hypertarget{ref-Sinha2015}{}%
9. \textbf{Sinha R}, \textbf{Chen J}, \textbf{Amir A}, \textbf{Vogtmann
E}, \textbf{Shi J}, \textbf{Inman KS}, \textbf{Flores R},
\textbf{Sampson J}, \textbf{Knight R}, \textbf{Chia N}. 2015. Collecting
fecal samples for microbiome analyses in epidemiology studies. Cancer
Epidemiology Biomarkers \& Prevention \textbf{25}:407--416.
doi:\href{https://doi.org/10.1158/1055-9965.epi-15-0951}{10.1158/1055-9965.epi-15-0951}.

\leavevmode\hypertarget{ref-Amir2017b}{}%
10. \textbf{Amir A}, \textbf{McDonald D}, \textbf{Navas-Molina JA},
\textbf{Debelius J}, \textbf{Morton JT}, \textbf{Hyde E},
\textbf{Robbins-Pianka A}, \textbf{Knight R}. 2017. Correcting for
microbial blooms in fecal samples during room-temperature shipping.
mSystems \textbf{2}:e00199--16.
doi:\href{https://doi.org/10.1128/msystems.00199-16}{10.1128/msystems.00199-16}.

\leavevmode\hypertarget{ref-Lauber2010}{}%
11. \textbf{Lauber CL}, \textbf{Zhou N}, \textbf{Gordon JI},
\textbf{Knight R}, \textbf{Fierer N}. 2010. Effect of storage conditions
on the assessment of bacterial community structure in soil and
human-associated samples. FEMS Microbiology Letters \textbf{307}:80--86.
doi:\href{https://doi.org/10.1111/j.1574-6968.2010.01965.x}{10.1111/j.1574-6968.2010.01965.x}.

\leavevmode\hypertarget{ref-Song2016}{}%
12. \textbf{Song SJ}, \textbf{Amir A}, \textbf{Metcalf JL},
\textbf{Amato KR}, \textbf{Xu ZZ}, \textbf{Humphrey G}, \textbf{Knight
R}. 2016. Preservation methods differ in fecal microbiome stability,
affecting suitability for field studies. mSystems \textbf{1}:e00021--16.
doi:\href{https://doi.org/10.1128/msystems.00021-16}{10.1128/msystems.00021-16}.

\leavevmode\hypertarget{ref-Costea2017}{}%
13. \textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa S},
\textbf{Pelletier E}, \textbf{Alberti A}, \textbf{Levenez F},
\textbf{Tramontano M}, \textbf{Driessen M}, \textbf{Hercog R},
\textbf{Jung F-E}, \textbf{Kultima JR}, \textbf{Hayward MR},
\textbf{Coelho LP}, \textbf{Allen-Vercoe E}, \textbf{Bertrand L},
\textbf{Blaut M}, \textbf{Brown JRM}, \textbf{Carton T},
\textbf{Cools-Portier S}, \textbf{Daigneault M}, \textbf{Derrien M},
\textbf{Druesne A}, \textbf{Vos WM de}, \textbf{Finlay BB},
\textbf{Flint HJ}, \textbf{Guarner F}, \textbf{Hattori M},
\textbf{Heilig H}, \textbf{Luna RA}, \textbf{Hylckama Vlieg J van},
\textbf{Junick J}, \textbf{Klymiuk I}, \textbf{Langella P},
\textbf{Chatelier EL}, \textbf{Mai V}, \textbf{Manichanh C},
\textbf{Martin JC}, \textbf{Mery C}, \textbf{Morita H}, \textbf{O'Toole
PW}, \textbf{Orvain C}, \textbf{Patil KR}, \textbf{Penders J},
\textbf{Persson S}, \textbf{Pons N}, \textbf{Popova M}, \textbf{Salonen
A}, \textbf{Saulnier D}, \textbf{Scott KP}, \textbf{Singh B},
\textbf{Slezak K}, \textbf{Veiga P}, \textbf{Versalovic J}, \textbf{Zhao
L}, \textbf{Zoetendal EG}, \textbf{Ehrlich SD}, \textbf{Dore J},
\textbf{Bork P}. 2017. Towards standards for human fecal sample
processing in metagenomic studies. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3960}{10.1038/nbt.3960}.

\leavevmode\hypertarget{ref-Haas2011}{}%
14. \textbf{Haas BJ}, \textbf{Gevers D}, \textbf{Earl AM},
\textbf{Feldgarden M}, \textbf{Ward DV}, \textbf{Giannoukos G},
\textbf{Ciulla D}, \textbf{Tabbaa D}, \textbf{Highlander SK},
\textbf{Sodergren E}, \textbf{Methe B}, \textbf{DeSantis TZ},
\textbf{Petrosino JF}, \textbf{Knight R}, \textbf{and BWB}. 2011.
Chimeric 16S rRNA sequence formation and detection in sanger and
454-pyrosequenced PCR amplicons. Genome Research \textbf{21}:494--504.
doi:\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.

\leavevmode\hypertarget{ref-Sinha2017}{}%
15. \textbf{Sinha R}, \textbf{Abu-Ali G}, \textbf{Vogtmann E},
\textbf{Fodor AA}, \textbf{Ren B}, \textbf{Amir A}, \textbf{Schwager E},
\textbf{Crabtree J}, \textbf{Ma S}, \textbf{Abnet CC}, \textbf{Knight
R}, \textbf{White O}, \textbf{Huttenhower C}. 2017. Assessment of
variation in microbial community amplicon sequencing by the microbiome
quality control (MBQC) project consortium. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3981}{10.1038/nbt.3981}.

\leavevmode\hypertarget{ref-Meisel2016}{}%
16. \textbf{Meisel JS}, \textbf{Hannigan GD}, \textbf{Tyldsley AS},
\textbf{SanMiguel AJ}, \textbf{Hodkinson BP}, \textbf{Zheng Q},
\textbf{Grice EA}. 2016. Skin microbiome surveys are strongly influenced
by experimental design. Journal of Investigative Dermatology
\textbf{136}:947--956.
doi:\href{https://doi.org/10.1016/j.jid.2016.01.016}{10.1016/j.jid.2016.01.016}.

\leavevmode\hypertarget{ref-Caporaso2010}{}%
17. \textbf{Caporaso JG}, \textbf{Lauber CL}, \textbf{Walters WA},
\textbf{Berg-Lyons D}, \textbf{Lozupone CA}, \textbf{Turnbaugh PJ},
\textbf{Fierer N}, \textbf{Knight R}. 2010. Global patterns of 16S rRNA
diversity at a depth of millions of sequences per sample. Proceedings of
the National Academy of Sciences \textbf{108}:4516--4522.
doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.

\leavevmode\hypertarget{ref-Kozich2013}{}%
18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\leavevmode\hypertarget{ref-Schloss2011}{}%
19. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
Reducing the effects of PCR amplification and sequencing artifacts on
16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.

\leavevmode\hypertarget{ref-Bokulich2012}{}%
20. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
DA}, \textbf{Caporaso JG}. 2012. Quality-filtering vastly improves
diversity estimates from illumina amplicon sequencing. Nature Methods
\textbf{10}:57--59.
doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.

\leavevmode\hypertarget{ref-Parada2015}{}%
21. \textbf{Parada AE}, \textbf{Needham DM}, \textbf{Fuhrman JA}. 2015.
Every base matters: Assessing small subunit rRNA primers for marine
microbiomes with mock communities, time series and global field samples.
Environmental Microbiology \textbf{18}:1403--1414.
doi:\href{https://doi.org/10.1111/1462-2920.13023}{10.1111/1462-2920.13023}.

\leavevmode\hypertarget{ref-Wang1996}{}%
22. \textbf{Wang GCY}, \textbf{Wang Y}. 1996. The frequency of chimeric
molecules as a consequence of PCR co-amplification of 16S rRNA genes
from different bacterial species. Microbiology \textbf{142}:1107--1114.
doi:\href{https://doi.org/10.1099/13500872-142-5-1107}{10.1099/13500872-142-5-1107}.
